---
title: "Crab Age Prediction"
author: "Trevor Okinda"
date: "2025"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Crab Age Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/sidhus/crab-age-prediction\>*

### Reference:

*\<Sidhu, G. S. (2021). Crab Age Prediction [Data set]. Kaggle. https://www.kaggle.com/datasets/sidhus/crab-age-prediction\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))
## Load dataset
```{r Load dataset}
# Load dataset
crab_data <- read.csv("CrabAgePrediction.csv", colClasses = c(
  Sex = "factor",
  Length = "numeric",
  Diameter = "numeric",
  Height = "numeric",
  Weight = "numeric",
  Shucked_Weight = "numeric",
  Viscera_Weight = "numeric",
  Shell_Weight = "numeric",
  Age = "integer"
))

# Display the structure of the dataset
str(crab_data)

# View the first few rows of the dataset
head(crab_data)

# View the dataset in a separate viewer window
View(crab_data)
```

## Measures of Frequency
```{r MOF}
# Load necessary libraries
library(ggplot2)

# Frequency distribution of categorical variable 'Sex'
sex_frequency <- table(crab_data$Sex)
print(sex_frequency)

# Plot frequency distribution of 'Sex'
ggplot(crab_data, aes(x = Sex)) + 
  geom_bar(fill = "lightblue", color = "black") +
  labs(title = "Frequency Distribution of Sex", x = "Sex", y = "Frequency")
```

## Measures of Central Tendency
```{r MOCT}
# Mean of numeric variables
mean_length <- mean(crab_data$Length, na.rm = TRUE)
mean_diameter <- mean(crab_data$Diameter, na.rm = TRUE)
mean_height <- mean(crab_data$Height, na.rm = TRUE)
mean_weight <- mean(crab_data$Weight, na.rm = TRUE)
mean_shucked_weight <- mean(crab_data$Shucked_Weight, na.rm = TRUE)
mean_viscera_weight <- mean(crab_data$Viscera_Weight, na.rm = TRUE)
mean_shell_weight <- mean(crab_data$Shell_Weight, na.rm = TRUE)
mean_age <- mean(crab_data$Age, na.rm = TRUE)

print(c(mean_length, mean_diameter, mean_height, mean_weight, mean_shucked_weight, mean_viscera_weight, mean_shell_weight, mean_age))

# Median of numeric variables
median_length <- median(crab_data$Length, na.rm = TRUE)
median_diameter <- median(crab_data$Diameter, na.rm = TRUE)
median_height <- median(crab_data$Height, na.rm = TRUE)
median_weight <- median(crab_data$Weight, na.rm = TRUE)
median_shucked_weight <- median(crab_data$Shucked_Weight, na.rm = TRUE)
median_viscera_weight <- median(crab_data$Viscera_Weight, na.rm = TRUE)
median_shell_weight <- median(crab_data$Shell_Weight, na.rm = TRUE)
median_age <- median(crab_data$Age, na.rm = TRUE)

print(c(median_length, median_diameter, median_height, median_weight, median_shucked_weight, median_viscera_weight, median_shell_weight, median_age))

```

## Measures of Distribution
```{r MOD}
# Standard deviation of numeric variables
sd_length <- sd(crab_data$Length, na.rm = TRUE)
sd_diameter <- sd(crab_data$Diameter, na.rm = TRUE)
sd_height <- sd(crab_data$Height, na.rm = TRUE)
sd_weight <- sd(crab_data$Weight, na.rm = TRUE)
sd_shucked_weight <- sd(crab_data$Shucked_Weight, na.rm = TRUE)
sd_viscera_weight <- sd(crab_data$Viscera_Weight, na.rm = TRUE)
sd_shell_weight <- sd(crab_data$Shell_Weight, na.rm = TRUE)
sd_age <- sd(crab_data$Age, na.rm = TRUE)

print(c(sd_length, sd_diameter, sd_height, sd_weight, sd_shucked_weight, sd_viscera_weight, sd_shell_weight, sd_age))

# Plot distribution of 'Length'
ggplot(crab_data, aes(x = Length)) + 
  geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Length", x = "Length", y = "Frequency")
```

## Measures of Relationship
```{r MOR}
# Correlation matrix of numeric variables
cor_matrix <- cor(crab_data[, c("Length", "Diameter", "Height", "Weight", "Shucked_Weight", "Viscera_Weight", "Shell_Weight", "Age")], use = "complete.obs")
print(cor_matrix)

# Plot correlation matrix
library(corrplot)
corrplot(cor_matrix, method = "circle")

# Scatter plot of Length vs Age
ggplot(crab_data, aes(x = Length, y = Age)) + 
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Length vs Age", x = "Length", y = "Age")

```

## ANOVA
```{r ANOVA}
# Load necessary library
library(ggplot2)
library(dplyr)
library(car)

# ANOVA: Weight by Sex
anova_result <- aov(Weight ~ Sex, data = crab_data)
summary(anova_result)

# Post-hoc test if ANOVA is significant
if (summary(anova_result)[[1]][["Pr(>F)"]][1] < 0.05) {
  print("ANOVA is significant, performing post-hoc test (Tukey HSD)...")
  tukey_result <- TukeyHSD(anova_result)
  print(tukey_result)
} else {
  print("ANOVA is not significant, no need for post-hoc test.")
}

# Boxplot to visualize differences in Weight by Sex
ggplot(crab_data, aes(x = Sex, y = Weight)) + 
  geom_boxplot(fill = 'orange', color = 'black') +
  labs(title = "Boxplot of Weight by Sex", x = "Sex", y = "Weight")
```

## Plots
```{r Plots}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Histogram for Length
ggplot(crab_data, aes(x = Length)) + 
  geom_histogram(binwidth = 0.1, fill = 'blue', color = 'black') +
  labs(title = "Histogram of Length", x = "Length", y = "Frequency")

# Density plot for Weight
ggplot(crab_data, aes(x = Weight)) + 
  geom_density(fill = 'green', alpha = 0.5) +
  labs(title = "Density Plot of Weight", x = "Weight", y = "Density")

# Bar plot for categorical variable (Sex)
ggplot(crab_data, aes(x = Sex)) + 
  geom_bar(fill = 'purple', color = 'black') +
  labs(title = "Bar Plot of Sex", x = "Sex", y = "Count")

# Scatter plot between Length and Weight colored by Sex
ggplot(crab_data, aes(x = Length, y = Weight, color = Sex)) + 
  geom_point() + 
  labs(title = "Scatter Plot of Length vs Weight", x = "Length", y = "Weight")

# Pair plot for numeric variables
pairs(crab_data[, sapply(crab_data, is.numeric)], main = "Pairwise Scatter Plots", 
      pch = 19, col = crab_data$Sex)

# Enhanced pair plot using GGally package
library(GGally)
ggpairs(crab_data, aes(color = Sex, alpha = 0.5))

# Box plot for Length by Sex
ggplot(crab_data, aes(x = Sex, y = Length)) + 
  geom_boxplot(fill = 'orange', color = 'black') +
  labs(title = "Box Plot of Length by Sex", x = "Sex", y = "Length")

# Violin plot for Weight by Sex
ggplot(crab_data, aes(x = Sex, y = Weight)) + 
  geom_violin(fill = 'cyan', color = 'black') +
  labs(title = "Violin Plot of Weight by Sex", x = "Sex", y = "Weight")

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}

```

