---
title: "Crab Age Prediction"
author: "Trevor Okinda"
date: "2025"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Crab Age Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/sidhus/crab-age-prediction\>*

### Reference:

*\<Sidhu, G. S. (2021). Crab Age Prediction [Data set]. Kaggle. https://www.kaggle.com/datasets/sidhus/crab-age-prediction\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))
## Load dataset
```{r Load dataset}
# Load dataset
crab_data <- read.csv("CrabAgePrediction.csv", colClasses = c(
  Sex = "factor",
  Length = "numeric",
  Diameter = "numeric",
  Height = "numeric",
  Weight = "numeric",
  Shucked_Weight = "numeric",
  Viscera_Weight = "numeric",
  Shell_Weight = "numeric",
  Age = "integer"
))

# Display the structure of the dataset
str(crab_data)

# View the first few rows of the dataset
head(crab_data)

# View the dataset in a separate viewer window
View(crab_data)
```

## Measures of Frequency
```{r MOF}
# Load necessary libraries
library(ggplot2)

# Frequency distribution of categorical variable 'Sex'
sex_frequency <- table(crab_data$Sex)
print(sex_frequency)

# Plot frequency distribution of 'Sex'
ggplot(crab_data, aes(x = Sex)) + 
  geom_bar(fill = "lightblue", color = "black") +
  labs(title = "Frequency Distribution of Sex", x = "Sex", y = "Frequency")
```

## Measures of Central Tendency
```{r MOCT}
# Mean of numeric variables
mean_length <- mean(crab_data$Length, na.rm = TRUE)
mean_diameter <- mean(crab_data$Diameter, na.rm = TRUE)
mean_height <- mean(crab_data$Height, na.rm = TRUE)
mean_weight <- mean(crab_data$Weight, na.rm = TRUE)
mean_shucked_weight <- mean(crab_data$Shucked_Weight, na.rm = TRUE)
mean_viscera_weight <- mean(crab_data$Viscera_Weight, na.rm = TRUE)
mean_shell_weight <- mean(crab_data$Shell_Weight, na.rm = TRUE)
mean_age <- mean(crab_data$Age, na.rm = TRUE)

print(c(mean_length, mean_diameter, mean_height, mean_weight, mean_shucked_weight, mean_viscera_weight, mean_shell_weight, mean_age))

# Median of numeric variables
median_length <- median(crab_data$Length, na.rm = TRUE)
median_diameter <- median(crab_data$Diameter, na.rm = TRUE)
median_height <- median(crab_data$Height, na.rm = TRUE)
median_weight <- median(crab_data$Weight, na.rm = TRUE)
median_shucked_weight <- median(crab_data$Shucked_Weight, na.rm = TRUE)
median_viscera_weight <- median(crab_data$Viscera_Weight, na.rm = TRUE)
median_shell_weight <- median(crab_data$Shell_Weight, na.rm = TRUE)
median_age <- median(crab_data$Age, na.rm = TRUE)

print(c(median_length, median_diameter, median_height, median_weight, median_shucked_weight, median_viscera_weight, median_shell_weight, median_age))

```

## Measures of Distribution
```{r MOD}
# Standard deviation of numeric variables
sd_length <- sd(crab_data$Length, na.rm = TRUE)
sd_diameter <- sd(crab_data$Diameter, na.rm = TRUE)
sd_height <- sd(crab_data$Height, na.rm = TRUE)
sd_weight <- sd(crab_data$Weight, na.rm = TRUE)
sd_shucked_weight <- sd(crab_data$Shucked_Weight, na.rm = TRUE)
sd_viscera_weight <- sd(crab_data$Viscera_Weight, na.rm = TRUE)
sd_shell_weight <- sd(crab_data$Shell_Weight, na.rm = TRUE)
sd_age <- sd(crab_data$Age, na.rm = TRUE)

print(c(sd_length, sd_diameter, sd_height, sd_weight, sd_shucked_weight, sd_viscera_weight, sd_shell_weight, sd_age))

# Plot distribution of 'Length'
ggplot(crab_data, aes(x = Length)) + 
  geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Length", x = "Length", y = "Frequency")
```

## Measures of Relationship
```{r MOR}
# Correlation matrix of numeric variables
cor_matrix <- cor(crab_data[, c("Length", "Diameter", "Height", "Weight", "Shucked_Weight", "Viscera_Weight", "Shell_Weight", "Age")], use = "complete.obs")
print(cor_matrix)

# Plot correlation matrix
library(corrplot)
corrplot(cor_matrix, method = "circle")

# Scatter plot of Length vs Age
ggplot(crab_data, aes(x = Length, y = Age)) + 
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Length vs Age", x = "Length", y = "Age")

```

## ANOVA
```{r ANOVA}
# Load necessary library
library(ggplot2)
library(dplyr)
library(car)

# ANOVA: Weight by Sex
anova_result <- aov(Weight ~ Sex, data = crab_data)
summary(anova_result)

# Post-hoc test if ANOVA is significant
if (summary(anova_result)[[1]][["Pr(>F)"]][1] < 0.05) {
  print("ANOVA is significant, performing post-hoc test (Tukey HSD)...")
  tukey_result <- TukeyHSD(anova_result)
  print(tukey_result)
} else {
  print("ANOVA is not significant, no need for post-hoc test.")
}

# Boxplot to visualize differences in Weight by Sex
ggplot(crab_data, aes(x = Sex, y = Weight)) + 
  geom_boxplot(fill = 'orange', color = 'black') +
  labs(title = "Boxplot of Weight by Sex", x = "Sex", y = "Weight")
```

## Plots
```{r Plots}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Histogram for Length
ggplot(crab_data, aes(x = Length)) + 
  geom_histogram(binwidth = 0.1, fill = 'blue', color = 'black') +
  labs(title = "Histogram of Length", x = "Length", y = "Frequency")

# Density plot for Weight
ggplot(crab_data, aes(x = Weight)) + 
  geom_density(fill = 'green', alpha = 0.5) +
  labs(title = "Density Plot of Weight", x = "Weight", y = "Density")

# Bar plot for categorical variable (Sex)
ggplot(crab_data, aes(x = Sex)) + 
  geom_bar(fill = 'purple', color = 'black') +
  labs(title = "Bar Plot of Sex", x = "Sex", y = "Count")

# Scatter plot between Length and Weight colored by Sex
ggplot(crab_data, aes(x = Length, y = Weight, color = Sex)) + 
  geom_point() + 
  labs(title = "Scatter Plot of Length vs Weight", x = "Length", y = "Weight")

# Pair plot for numeric variables
pairs(crab_data[, sapply(crab_data, is.numeric)], main = "Pairwise Scatter Plots", 
      pch = 19, col = crab_data$Sex)

# Enhanced pair plot using GGally package
library(GGally)
ggpairs(crab_data, aes(color = Sex, alpha = 0.5))

# Box plot for Length by Sex
ggplot(crab_data, aes(x = Sex, y = Length)) + 
  geom_boxplot(fill = 'orange', color = 'black') +
  labs(title = "Box Plot of Length by Sex", x = "Sex", y = "Length")

# Violin plot for Weight by Sex
ggplot(crab_data, aes(x = Sex, y = Weight)) + 
  geom_violin(fill = 'cyan', color = 'black') +
  labs(title = "Violin Plot of Weight by Sex", x = "Sex", y = "Weight")

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in each column
missing_values <- sapply(crab_data, function(x) sum(is.na(x)))
print(missing_values)

# Display the columns with missing values and their counts
missing_values_df <- data.frame(Column = names(missing_values), MissingValues = missing_values)
print(missing_values_df)

# Visualize missing values using the VIM package
library(VIM)
aggr(crab_data, col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, labels = names(crab_data), cex.axis = 0.7, gap = 3, ylab = c("Missing data", "Pattern"))
```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load necessary libraries
library(caret)
library(dplyr)

# Split the data into training (80%) and test (20%) sets
set.seed(123)
trainIndex <- createDataPartition(crab_data$Age, p = 0.8, list = FALSE)
train_data <- crab_data[trainIndex,]
test_data <- crab_data[-trainIndex,]

# Display the size of training and test sets
cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")
```

## Bootstrapping
```{r Bootstrapping}
# Load necessary libraries
library(boot)

# Define a function to calculate the mean of Weight
mean_weight <- function(data, indices) {
  sample_data <- data[indices, ]
  return(mean(sample_data$Weight))
}

# Perform bootstrapping
set.seed(123)
boot_results <- boot(data = crab_data, statistic = mean_weight, R = 1000)

# Display bootstrapping results
print(boot_results)

# Plot the bootstrapping results
plot(boot_results)
```

## Cross-validation
```{r Cross-validation}
# Load necessary libraries
library(caret)

# Define a control function for k-fold cross-validation
control <- trainControl(method = "cv", number = 10)

# Train a linear regression model using k-fold cross-validation
set.seed(123)
cv_model <- train(Age ~ Length + Diameter + Height + Weight + Shucked_Weight + Viscera_Weight + Shell_Weight,
                  data = train_data,
                  method = "lm",
                  trControl = control)

# Display the cross-validation results
print(cv_model)

# Predict on the test set
predictions <- predict(cv_model, newdata = test_data)

# Calculate and display performance metrics
mse <- mean((predictions - test_data$Age)^2)
rmse <- sqrt(mse)
cat("Mean Squared Error:", mse, "\n")
cat("Root Mean Squared Error:", rmse, "\n")
```

## Train Different Models
```{r Train Different models}
# Load necessary libraries
library(caret)
library(rpart)
library(randomForest)

# Train a linear regression model
lm_model <- train(Age ~ Length + Diameter + Height + Weight + Shucked_Weight + Viscera_Weight + Shell_Weight,
                  data = train_data,
                  method = "lm")

# Display the linear regression model summary
print(summary(lm_model$finalModel))

# Predict on the test set
lm_predictions <- predict(lm_model, newdata = test_data)

# Calculate performance metrics
lm_mse <- mean((lm_predictions - test_data$Age)^2)
lm_rmse <- sqrt(lm_mse)
cat("Linear Regression - Mean Squared Error:", lm_mse, "\n")
cat("Linear Regression - Root Mean Squared Error:", lm_rmse, "\n")

# Train a decision tree model
dt_model <- train(Age ~ Length + Diameter + Height + Weight + Shucked_Weight + Viscera_Weight + Shell_Weight,
                  data = train_data,
                  method = "rpart")

# Display the decision tree model summary
print(dt_model$finalModel)

# Plot the decision tree
plot(dt_model$finalModel)
text(dt_model$finalModel, use.n = TRUE)

# Predict on the test set
dt_predictions <- predict(dt_model, newdata = test_data)

# Calculate performance metrics
dt_mse <- mean((dt_predictions - test_data$Age)^2)
dt_rmse <- sqrt(dt_mse)
cat("Decision Tree - Mean Squared Error:", dt_mse, "\n")
cat("Decision Tree - Root Mean Squared Error:", dt_rmse, "\n")

# Train a random forest model
set.seed(123)
rf_model <- randomForest(Age ~ Length + Diameter + Height + Weight + Shucked_Weight + Viscera_Weight + Shell_Weight,
                         data = train_data, ntree = 500)

# Display the random forest model summary
print(rf_model)

# Predict on the test set
rf_predictions <- predict(rf_model, newdata = test_data)

# Calculate performance metrics
rf_mse <- mean((rf_predictions - test_data$Age)^2)
rf_rmse <- sqrt(rf_mse)
cat("Random Forest - Mean Squared Error:", rf_mse, "\n")
cat("Random Forest - Root Mean Squared Error:", rf_rmse, "\n")
```

## Performance Metrics
```{r Models performance w resamples}
# Load necessary libraries
library(caret)
library(dplyr)
library(reshape2)

# Define training control
train_control <- trainControl(method = "cv", number = 10)

# Define models
models <- list(
  "Linear Regression" = list(model = "lm", method = "lm"),
  "Decision Tree" = list(model = "rpart", method = "rpart"),
  "Random Forest" = list(model = "rf", method = "rf")
)

# Train and evaluate models
results <- lapply(models, function(model) {
  set.seed(123)
  train(Age ~ Length + Diameter + Height + Weight + Shucked_Weight + Viscera_Weight + Shell_Weight,
        data = crab_data,
        method = model$method,
        trControl = train_control)
})

# Extract performance metrics
performance <- lapply(results, function(model) {
  resamples <- model$resample
  resamples$Model <- rownames(resamples)
  return(resamples)
})

# Combine results into a single data frame
performance_df <- do.call(rbind, performance)

# Reshape data for visualization
performance_melted <- melt(performance_df, id.vars = "Model", variable.name = "Metric", value.name = "Value")

# Plot model performance
ggplot(performance_melted, aes(x = Model, y = Value, fill = Model)) +
  geom_boxplot() +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(title = "Model Performance Comparison", y = "Value")
```

# Saving Model
```{r Saving Model}
# Saving the linear regression model
saveRDS(lm_model, "./models/saved_lm_model.rds")

# Load the saved model
loaded_lm_model <- readRDS("./models/saved_lm_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  Length = 1.5,
  Diameter = 1.2,
  Height = 0.4,
  Weight = 30,
  Shucked_Weight = 15,
  Viscera_Weight = 7,
  Shell_Weight = 8
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_lm_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```


